---
title: "Hemoglobin distributions and trends"
author: "Timo Asikainen"
date: "2025-02-20" 
output: html_document
---

## How to run the code

### Parameter settings

Some parameters must be set for the code to run correctly. The parameters 
are saved in the **param** list. The parameters are set in the **parameters** block
below. The parameters are:

- wd: The working directory to be used. This is set by Rstudio to either the **src**
or the root folder of the project, e.g., "C:/git_repot/first-time-donor-hb".
The correct setting is the project root. The code in the block checks if the 
**src** folder has been selected, and changes the folder to project root if yes.
- result.file: where the anonymised results will be exported. By default this
is in the **results** subfolder in file named **exported-data.xlsx**. This is computed
by the code and needs not be changed.
- country: two-letter identifier for the country, e.g., FI. **Should be changed.**
- age.minimum, age.maximum: The minimum and maximum ages, respectively, at the
time of donation that are included in the exported data.
- hb.minimum, hb.maximum: Similarly as above, but for Hb values. The default
values are -Inf and Inf, respectively, meaning that all Hb non-NA/NULL Hb values 
are included. 
- include.na: Should donations with Hb value unavailable (NA) included or not.
Default is yes (**TRUE**).
- minimum.group.size: The smallest group size (*n*) that is exported.
- donation.type.keys: Which values in **donationdata$donation$BloodDonationTypeKey**
are included in the analysis. The idea is to consider full blood donations, 
either successful or deferred cases.
- hb.decimals: The number of decimals in Hb values that should be used.
- donationdata: Path to the .Rdata file containing the data to be used. This is
hard-coded by default to **donationdata.Rdata** in the working directory.
- units: HB units to be used: Shold be one of: mmol/L, or g/L or g/dL.
- cutoff.male, cutoff.female: The required mimimum values to donate for males 
and females, respectively. These **must be set** and expressed in **units**.
- data.set: Which data sets should be summarised and exported. The default is
c('donation0','donation.r','simple'), which correspond to (1) first time donations 
(first mention of releaseID in the **donation** data set); (2) repeat donations, 
i.e., other than the first mention; and (3) all donations. The primary interest 
lies in the first donations, but other data sets are included by default for 
completeness and to enable additional analyses.
- extractHour: A function to extract the hour from **donationdata**. The default 
provided works when the time information is included in **DonationTimeDTTM** 
column in **donationdata$donation** in POSIXct format. The function
can be adjusted accordingnly.
- donation.cols, donor.cols: Some columns maybe have multiple sources in
donationdata. These settings are vectors that specify the columns that are to be
copied from donationdata$donation and donationdata$donor, respectively. 
The default values are **c()** (no columns) and **c('DateOfBirth','Sex','BloodGroup'),**
again, respectively. These values correspond to the **data-description.xlsx** file.
That is, by default the date of birth, sex and blood group
are copied from the donor data to the working data set. The values can be copied
from donor.cols to donation.cols to alter the behaviour. In addition, column(s) 
that is used to compute the hour of the donation with the day, should be added
to **donation.cols**, as is done country-specifically under **NL** and **FI**.

## Exporting the data

To export the data, code should be run from the start of this file 
(** main-with-export.Rmd**) up to and including the chunk **export-data**. 
This includes setting the parameters (described) above, loading and processing 
the data, extracting anonymous statistics and writing these to the 
**exported-data.xlsx** file as described above.

The statistics are anonymised by removing any groups with size smaller then 
the parameter **minimum.group.size** (default 5). For Hb, small and large Hb
values with fewer observations than the limit are pooled together and inxluded
in the exported data using specific placeholders (-1000000 and 1000000). 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-packages}
# install.packages('moments')
library(dplyr)
library(lubridate)
library(tidyr)
library(moments)
library(openxlsx)
```

```{r parameters}
param=list()
param$wd = getwd()

# getwd() might behave differently depending on the environment where it is run (console vs. Rmd),
# therefore checking if working directory is set to the src folder and moving up if yes.
if (grepl('[/\\]src[/\\]?',param$wd)) {
   param$wd = sub('[/\\]src([/\\]?)$','\\1',param$wd)
}

dir.create(file.path(param$wd,"results"),showWarnings = FALSE)
dir.create(file.path(param$wd,"log"),showWarnings = FALSE)
param$result.file = file.path(param$wd,"results","exported-data.xlsx")

# param$country = 'NL'
param$country = 'FI'

param$include.na = TRUE
param$minimum.group.size = 5
param$donation.type.keys = NULL

param$age.minimum = 16
param$age.maximum = 75
param$hb.minimum = -Inf
param$hb.maximum = Inf

# This is the default file location. If you are using a country-specific file
# name or path, please specify it below under country-specific parameters
param$donationdata = file.path(param$wd,'donationdata.Rdata')

# The number of intended decimals in Hb values
# Typically 0 for g/L, 1 for g/dL and mmol/L
param$hb.decimals = NA

param$donor.cols = c('DateOfBirth','Sex','BloodGroup')
param$donation.cols = c()

# conversion factor from mmol/L to g/L: cf = 10 / 0.6206, see import-data.Rmd for details

# always return 0 (if no hour is available)
# nb! In order to extract hourly data, this function should be redefined as a
# country-specific parameter
param$extractHour = function(simple) {return(0)} 

# Country-specific parameter settings
if (param$country == 'NL') {
  param$donationdata = '~/data/data_timo.rdata'
  param$units = 'mmol/L' # one of: mmol/L, or g/L or g/dL
  param$cutoff.male = 8.4 # will need to convert these to correct units
  param$cutoff.female = 7.8
  param$data.sets = c('donation0','donation.r','simple')
  # for NL the correct Sex and DateofBirth is in donations... sorry this is a bit cumbersome
  param$hb.minimum = 1
  param$hb.maximum = 20
  param$hb.decimals = 1 #just to be sure
  # MP: Funky hour extraction for NL...
  param$extractHour = function(simple){ 
    return(ifelse(is.na(simple$DonationTimeStart), NA_integer_, as.integer(substr(simple$DonationTimeStart, 1, 2))))
  }
  param$donor.cols = c('BloodGroup')
  param$donation.cols = c('Sex','DateOfBirth','DonationTimeStart')
} else if (param$country == 'FI') {
  param$donationdata = 'C:/git_repot/DATA/donationdata.fortimo.rdata'
  param$units = 'g/L' # one of: mmol/L or g/L or g/dL
  param$cutoff.male = 135
  param$cutoff.female = 125
  param$hb.minimum = 50
  param$hb.maximum = 250
  param$data.sets = c('donation0','donation.r','simple')
  param$donation.type.keys = c('Whole Blood (K)','No Donation (E)','VisitNoDonation')
  param$hb.decimals = 0
  param$extractHour = function(simple) {
    return(as.integer(format(simple$DonationTimeDTTM,'%H')))
  }
  param$donor.cols = c('DateOfBirth','Sex','BloodGroup')
  param$donation.cols = c('DonationTimeDTTM')
} 

# automated reasoning in case param$hb.decimals is not set
if (is.na(param$hb.decimals)) {
  if (param$units %in% c('mmol/L','g/dL'))
    hb.decimals = 1
  else
    hb.decimals = 0
}
```

```{r load-data}
# The data is expected to be in the same for as in the long-term forecasts project, except for
# the column 'Hb' added in the donation table. The units used in Finland is g/l.
load(param$donationdata)
```


```{r load-and-process-data}
if (is.null(param$donation.type.keys)) 
  param$donation.type.keys = unique(donationdata$donation$BloodDonationTypeKey)

simple = donationdata$donation[, c("releaseID", "BloodDonationTypeKey", "DonationDate", 
                  "DonationPlaceType", "DonationPlaceCode", "Hb", param$donation.cols)] %>%
  filter(param$include.na || !is.na(Hb)) %>%
  filter(BloodDonationTypeKey %in% param$donation.type.keys) %>%
  # VisitNoDonation is used in Finland during the recent years instead of No Donation (E)
  left_join(donationdata$donor[,c('releaseID',param$donor.cols)],by='releaseID')
simple$age= as.numeric(difftime(simple$DonationDate,simple$DateOfBirth),unit="weeks")/52.25
simple$month = as.factor(month(simple$DonationDate))
simple$year = year(simple$DonationDate)
simple$year.factor = as.factor(simple$year)

# 2025-05-15 New line to convert the new datetime column to integer for hour)
simple$hour = param$extractHour(simple)

simple = simple %>%
  filter(age >= param$age.minimum, age <= param$age.maximum) %>%
  filter(Hb >= param$hb.minimum, Hb <= param$hb.maximum) %>%
  mutate(Hb = round(Hb,param$hb.decimals)) %>%
  filter(Sex %in% c('Female','Male')) 

simple$limit = param$cutoff.male
simple$limit[simple$Sex=='Female'] = param$cutoff.female
simple$deferred=0
simple$deferred[simple$Hb<simple$limit]=1

#Filter nan sex
simple <- simple %>% filter(!is.na(Sex))

#For NL should take only BloodDonationTypeKey == "N" for first donation
# date of first donation
#The filter(param$country != NL || BloodDonationTypeKey == "New") gave an error for me so changed to this
if (param$country == "NL") {
  date0 = simple %>%
    filter(BloodDonationTypeKey == "New") %>%
    group_by(releaseID) %>%
    summarise(date0=min(DonationDate),.groups='drop')
  } else {
    date0 = simple %>%
    group_by(releaseID) %>%
    summarise(date0=min(DonationDate),.groups='drop')
}

# first-time donations
donation0 = simple %>%
  inner_join(date0,join_by(x$DonationDate==y$date0,releaseID))

# repeat donations; sent for sentinel
donation.r = simple %>%
  left_join(cbind(date0,sent=1),join_by(x$DonationDate==y$date0,releaseID)) %>%
  filter(is.na(sent)) %>%
  select(-sent) %>%
  mutate(month=as.integer(month))

# donation0$year=year(donation0$DonationDate)
# donation0$year.factor=as.factor(donation0$year)
```

```{r prepare-for-export}
# Annual Hb distributions by Sex
data.sets= param$data.sets # c('donation0','donation.r','simple')
hb.freq = NULL
for (ds in data.sets) {
  freq0 = get(ds) %>%
    mutate(data.set=if(ds == 'simple') 'all' else ds) %>%
    mutate(month=as.integer(month)) %>%
    group_by(data.set,year,Sex,Hb) %>%
    summarise(n=n(),nas=sum(1*is.na(Hb)),deferred=sum(deferred,na.rm=TRUE),
              mean.hour=mean(hour,na.rm=TRUE),sd.hour=sd(hour,na.rm=TRUE),
              mean.month=mean(month,na.rm=TRUE),sd.month=sd(month,na.rm=TRUE),
              mean.age=mean(age,na.rm=TRUE),sd.age=sd(age,na.rm=TRUE),limit=min(limit),.groups='drop')
  hb.freq = if(is.null(hb.freq)) freq0 else rbind(hb.freq,freq0)
}

anonymizeDistribution = function(freq) {
  # freq = hb.freq
  hb.inf = 1e6
  wh.sub = which(freq$n < param$minimum.group.size & freq$Hb < freq$limit)
  wh.super = which(freq$n < param$minimum.group.size & freq$Hb > freq$limit)
  
  freq$Hb[wh.sub] = -hb.inf
  freq$Hb[wh.super] = hb.inf
  
  aggregated = freq[c(wh.sub,wh.super),] %>%
    mutate(sgn=sign(Hb)) %>%
    group_by(data.set,year,Sex,sgn) %>%
    summarise(Hb=min(sgn)*hb.inf,n2=sum(n),nas=sum(nas),deferred=sum(deferred),
              mean.age=sum(n*mean.age)/sum(n),sd.age=NA,
              mean.hour=sum(n*mean.hour)/sum(n),sd.hour=NA,
              mean.month=sum(n*mean.month)/sum(n),sd.month=NA,
              limit=min(limit),.groups='drop') %>%
    dplyr::select(-sgn) %>%
    rename(n=n2)

  freq=freq[-c(wh.sub,wh.super),]
  freq = rbind(freq,aggregated)
  freq = freq[freq$n >= param$minimum.group.size,]
  freq[with(freq,order(data.set,year,Sex,Hb)),]
}

hb.freq = anonymizeDistribution(hb.freq)

# Annual age distributions by Sex
data.sets=c('donation0','donation.r','simple')
age.freq = NULL
for (ds in data.sets) {
  freq0 = get(ds) %>%
    mutate(data.set=if(ds == 'simple') 'all' else ds,age=as.integer(age)) %>%
    group_by(data.set,year,Sex,age) %>%
    summarise(n=n(),nas=sum(1*is.na(Hb)),deferred=sum(deferred,na.rm=TRUE),mean.hb=mean(Hb,na.rm=TRUE),sd.hb=sd(Hb,na.rm=TRUE),.groups='drop') %>%
    filter(n >= param$minimum.group.size)
  age.freq = if(is.null(age.freq)) freq0 else rbind(age.freq,freq0)
}

# Month statistics
monthly = NULL
for (ds in data.sets) {
  monthly0 = get(ds) %>%
    mutate(data.set=ds) %>%
    group_by(data.set,year,Sex,month) %>%
    summarise(n=n(),nas=sum(1*is.na(Hb)),deferred=sum(deferred,na.rm=TRUE),mean=mean(Hb,na.rm=TRUE),sd=sd(Hb,na.rm=TRUE),
              mean.age=mean(age,na.rm=TRUE),sd.age=sd(age,na.rm=TRUE),nas=sum(1*is.na(Hb)),.groups='drop') %>%
    filter(n >= param$minimum.group.size) %>%
    arrange(year,Sex,month)
  monthly = if(is.null(monthly)) monthly0 else rbind(monthly,monthly0)
}

# Hourly statistics
hourly = NULL
for (ds in data.sets) {
  hourly0 = get(ds) %>%
    mutate(data.set=ds) %>%
    group_by(data.set,year,Sex,hour) %>%
    summarise(n=n(),nas=sum(1*is.na(Hb)),deferred=sum(deferred,na.rm=TRUE),mean=mean(Hb,na.rm=TRUE),sd=sd(Hb,na.rm=TRUE),
              mean.age=mean(age,na.rm=TRUE),sd.age=sd(age,na.rm=TRUE),nas=sum(1*is.na(Hb)),.groups='drop') %>%
    filter(n >= param$minimum.group.size) %>%
    arrange(year,Sex,hour)
  hourly = if(is.null(hourly)) hourly0 else rbind(hourly,hourly0)
}
```

Open questions:
- What to do with the garrison donations?
- What to do with the NA's?
- Should the lower parts be somehow masked: is there a risk of leaking personal data?

```{r export-data}
write.xlsx(list(parameters=data.frame(name=names(param),value=paste(param,sep=',')),
                annual.hb=hb.freq,annual.age=age.freq,montly.statistics=monthly,hourly.statistics=hourly),
           file=param$result.file)

# Remove the placeholder values for very small and very large Hb values and NA's
hb.freq = hb.freq[abs(hb.freq$Hb) < 1000 & !is.na(hb.freq$Hb),]
```

This is the end of the processing and export part.

```{r function-definitions}
plotColBySex = function(basics,col,ylab,xvar='year') {
  sex.col = data.frame(sex=c('Female','Male'),col=c('red','blue'))
  rownames(sex.col)=sex.col$sex
  
  par(mar=c(2,4,0,4))
  plot.cols=c(col)
  plot(basics[[xvar]],basics[[plot.cols[1]]],type='n',ylab=ylab)
  for (i in 1:nrow(sex.col)) {
    data1 = basics[basics$Sex==sex.col$sex[i],]
    for (col in plot.cols) {
      lines(data1[[xvar]],data1[[col]],col=sex.col$col[i],lwd=2)
      
      for (suf in c('low','hi')) {
        ci.var = paste0(col,'.',suf)
        if (ci.var %in% names(basics)) {
          lines(data1[[xvar]],data1[[ci.var]],lty='dotted',lwd=1,col=sex.col$col[i])
        }
      }
      
      frml=paste0(col,'~',xvar)
      m=lm(formula(frml),data=data1)
      sm=summary(m)
      abline(a=sm$coeff[1,1],b=sm$coeff[2,1],col=sex.col$col[i],lty='dashed',lwd=2)
      text(x=min(data1[[xvar]]),y=min(data1[[col]])+0.01,
           labels=paste0('b=',round(100*sm$coeff[2,1],5),'%,p=',round(sm$coeff[2,4],3)),adj=c(0))
    }
  }
}

# Some of these function definitions may be obsolete, should check and remove the unnecessary ones.
plotHbDist = function(data0,sex,cutoff,zoom=TRUE) {
  sex0=data0[data0$year>=2005&data0$Sex==sex & 
              !is.na(data0$Hb)&data0$BloodDonationTypeKey %in% c('Whole Blood (K)','No Donation (E)'),c('Hb')]  
  freq = sex0 %>%
    group_by(Hb) %>%
    summarise(n=n()) %>%
    mutate(prop=n/sum(n))

  limits = NULL
  if (zoom)
    limits = c(cutoff-5,cutoff+5)
  plot(prop~Hb,data=freq,type='l',lwd=2,xlim=limits)

  mean0=mean(sex0$Hb)
  sd0=sd(sex0$Hb)
  
  text(x=cutoff,y=0.03,labels=paste0('mean=',mean0,', sd=',sd0))
  
  ndist=dnorm(unique(freq$Hb),mean=mean0,sd=sd0)
  lines(unique(freq$Hb),ndist)
  abline(v=cutoff,col='blue')
  
  return(list(mean0=mean0,sd0=sd0))
}

estimateTrend = function(final) {
  final.longer = matToLonger(final)
  final.longer$interceptFemale=1
  final.longer$interceptFemale[final.longer$Sex=='Male']=0
  final.longer$interceptMale=1-final.longer$interceptFemale
  m=lm(value~0+interceptFemale+interceptMale+year:Sex,data=final.longer)
  print(summary(m))
  return(m)
}

matToLonger = function(mat) {
  df1=as.data.frame(pivot_longer(cbind(Sex=rownames(mat)[1:2],mat[1:2,]),colnames(mat),names_to='year'))
  df1$year=as.integer(df1$year)
  return(df1)
}

# This function is in active use, but it doesn't seem to refer the functions above
plotYearSex = function(m,relative=FALSE) {
  cofs = summary(m)$coeff
  conames=rownames(cofs)
  codf=data.frame(name=conames,estimate=cofs[,1])
  wh=grep('(Male|Female)',conames)
  codf$Sex[wh]=sub('.*(Male|Female).*','\\1',conames[wh])
  wh=grep('[0-9][0-9][0-9][0-9]',conames)
  codf$year[wh]=sub('[^0-9]+([0-9]+).*','\\1',conames[wh])
  codf=cbind(codf,confint(m))
  
  codf=codf[!is.na(codf$year),]
  pah=pivot_wider(codf,names_from='year',values_from='estimate',id_cols=c('Sex'))
  rn = as.vector(pah[,1])
  pah=as.data.frame(pah[,-1])
  rownames(pah)=rn$Sex
  
  means=apply(pah,1,mean)
  if (!relative)
    means=0*means
  
  pah=pah-means
  
  pah2=pivot_wider(codf,names_from='year',values_from='2.5 %',id_cols=c('Sex'))
  pah2=as.data.frame(pah2[,-1])
  pah2=pah2-means
  pah3=pivot_wider(codf,names_from='year',values_from='97.5 %',id_cols=c('Sex'))
  pah3=as.data.frame(pah3[,-1])
  pah3=pah3-means
  
  pah.all=rbind(pah,pah2,pah3)
  
  matplot(t(pah.all),type='l',lty=c(rep('solid',2),rep('dashed',4)),col=rep(c('red','blue'),2),axes=FALSE)
  axis(2)
  axis(side=1,at=1:ncol(pah.all),labels=colnames(pah.all))
  return(pah.all)
}
```

```{r distribution-analysis}
# This is the function that correct the kink in a distriution (data0)
# The function can be either called with data0 != NULL; data0 should then have the structure of simple, donation0 or donation.r
#   and can be a subset of these.
# *or* with freq != NULL, with the structure of hb.freq
# Nb! This function operates on a single distribution
rectifyDistribution = function (data0,sex=NULL,cutoff=NULL,plot=TRUE,freq=NULL) {
  if (is.null(cutoff)) {
    if (is.null(sex)) {
      if (is.null(data0)) {
        sex = unique(freq$Sex)[1]
      } else
        sex = unique(data0$Sex)[1]
    } 
    cutoff = if (sex=='Female') param$cutoff.female else param$cutoff.male
  }

  if (!is.null(data0)) {
    # sex0=data0[data0$Sex==sex&!is.na(data0$Hb)&data0$BloodDonationTypeKey %in% c('Whole Blood (K)','No Donation (E)'),c('Hb')]  
    freq = data0 %>%
      group_by(Hb) %>%
      summarise(n=n()) %>%
      mutate(prop=n/sum(n))
  } else {
    freq = freq %>%
      group_by(Hb) %>%
      summarise(n=sum(n),.groups='drop')
  }
  
  freq = freq[!is.na(freq$Hb),]
  
  if (!'prop' %in% colnames(freq)) {
    sum.n = sum(freq$n) # -coalesce(freq$nas,0))
    freq$prop = freq$n / sum.n
  }
  
  freq.mean = freq %>%
    filter(!is.na(Hb)) %>% # These are the summary rows for NA; leave them out at this point
    summarise(n2=sum(n),mean0=sum(Hb*(n))/n2,.groups='drop')
  
  freq.stats = freq %>%
    filter(!is.na(Hb)) %>% # These are the summary rows for NA; leave them out at this point
    mutate(dev=Hb-freq.mean$mean0) %>% # ,prop=n/n2) %>%
    group_by() %>%
    summarise(mean=min(freq.mean$mean0),n0=min(n),var=sum(prop*dev^2),sd=sqrt(var),
              skewness=sum(prop*(dev/sd)^3),kurtosis=sum(prop*(dev/sd)^4),.groups='drop')
  mean0= freq.stats$mean
  sd0 = freq.stats$sd

  Hb.values = sort(unique(freq$Hb))
  if (FALSE) {
    dx <- 1. #depends on units, but should now all be converted to g/L
  } else  {
    if (param$hb.decimals == 1){
      dx <- 0.1 #depends on units, but should now all be converted to g/L
    } else{
      dx = 1.
    }
  }
  ndist=dnorm(Hb.values,mean=mean0,sd=sd0)*dx
  
  # Plotting the unchanged distribution
  if (plot) {
    plot(prop~Hb,data=freq,type='l',lwd=2)
    lines(Hb.values,ndist)
  
    # A plot showing the difference between the distribution defined by data0 and the
    # normal distribution estimated from that data
    plot(freq$prop-ndist~Hb.values)
    abline(v=cutoff,col='blue')
    abline(h=c(-1,1)*0.0005,col='red')
  }

  # This is the part that 
  cnt = 0
  freq2 = freq
  # cover the case that there is no data at the cutoff value (should not occur with reasonable data volumes)
  # k = which(freq$Hb==cutoff)
  k = max(which(freq$Hb<=cutoff))
  if (length(k) == 0 || k == -Inf) 
    k = min(which(freq$Hb>=cutoff))
  while (TRUE) {
    diff = (freq2$prop-ndist)[1:(k-1)]
    wh = which(diff < -0.0005)
    if (length(wh) == 0)
      break
    wh0 = max(wh)
    ds0 = diff[wh0]
    
    diff.plus = (freq2$prop-ndist)[k:nrow(freq2)]
    wh = which(diff.plus > 0.0005)
    if (length(wh) == 0)
      break
    wh1 = min(wh)
    ds1 = diff.plus[wh1]
    
    to.adjust = min(-ds0,ds1)
    
    freq2$prop[wh0] = freq2$prop[wh0] + to.adjust
    freq2$prop[wh1+(k-1)] = freq2$prop[wh1+(k-1)] - to.adjust
    
    cnt = cnt + 1
    if (cnt > 1000) {
      print('max iterations exceeded (back-stop)')
      break
    }
  }

  mean1 = freq2 %>% 
      mutate(mom=Hb*prop) %>%
      summarise(mean=sum(mom))
  mean1=as.numeric(mean1)
  sd1 =  freq2 %>% 
      mutate(mom=(Hb-as.numeric(mean1))^2*prop) %>%
      summarise(sdx=sum(mom)) 
  sd1=sqrt(as.numeric(sd1))
  
  # The theoretical deferred proportion is computed here
  deferred.prop = pnorm(cutoff-0.5,mean1,sd1)

  if (plot) {
    plot(prop~Hb,data=freq2,type='l',lwd=3,xlim=NULL)
    lines(prop~Hb,data=freq2,col='green', lwd=2)
    ndist2=dnorm(Hb.values,mean=mean1,sd=sd1)*dx
    lines(Hb.values,ndist2,col='red',lty='dotted',lwd=2)
    abline(v=cutoff,col='blue',lty='dashed')
    rect(mean1-5,0,mean1+5,0.001,col='pink',lwd=2)
    abline(v=mean1,lty='dotted',lwd=3)
    
    plot(freq2$prop-ndist2~Hb.values)
    abline(v=cutoff,col='blue',lty='dashed')
    abline(h=c(-1,1)*0.0005,col='red',lty='dashed')
  }

  return(list(dist=freq2,params=list(mean=mean1,sd=sd1,deferred.prop=deferred.prop)))
}
```


```{r}
# Example of rectifying the distribution
freq <- hb.freq[hb.freq$Sex=='Female', ]
rectifyDistribution(NULL,'Female',param$cutoff.female,freq=freq)
# dist.male = rectifyDistribution(simple,'Male',cutoff)
# dist.female = rectifyDistribution(simple,'Female',cutoff)
```


```{r getStats-function}
# This function produces statistics of value.col (Hb or age) based on groups defined by group.col (can be year or age) and sex
# If age.freq is provided, the proportions are included in the results (relevant when value.col == 'Hb')
getStats = function(freq,value.col='Hb',group.col='year',n.filter=100,age.freq=NULL,age.breaks=c(-Inf,24,40,Inf),age.labels=c('young','mid','old')) {
  freq = data.frame(freq)
  
  if (is.null(n.filter) || is.na(n.filter))
    n.filter = 0
  
  wh = min(which(colnames(freq)==value.col))
  colnames(freq)[wh]='value'
  wh.group = min(which(colnames(freq)==group.col))
  colnames(freq)[wh.group]='group'

  new.basics.mean = freq %>%
    # filter(is.null(data.set) || data.set==data.set) %>%
    filter(!is.na(value)) %>% # These are the summary rows for NA; leave them out at this point
    group_by(group,Sex) %>%
    summarise(n2=sum(n-nas),deferred.prop=sum(deferred/n2),mean0=sum(value*(n-nas))/n2,.groups='drop')
  
  new.basics = freq %>%
    filter(!is.na(value)) %>% # These are the summary rows for NA; leave them out at this point
    inner_join(new.basics.mean,join_by(group,Sex)) %>%
    mutate(dev=value-mean0,prop=n/n2) %>%
    group_by(group,Sex) %>%
    summarise(mean=min(mean0),n0=min(n),n=min(n2),deferred.prop=min(deferred.prop),var=sum(prop*dev^2),sd=sqrt(var),
              skewness=sum(prop*(dev/sd)^3),kurtosis=sum(prop*(dev/sd)^4),.groups='drop') %>%
    mutate(sd.low=sqrt((n-1)*var/qchisq(c(.975), n-1))) %>%
    mutate(sd.hi =sqrt((n-1)*var/qchisq(c(.0275), n-1))) %>%
    mutate(mean.low=mean-qnorm(0.025)*sd/sqrt(n)) %>%
    mutate(mean.hi =mean-qnorm(0.975)*sd/sqrt(n)) %>%
    filter(n>n.filter) %>%
    data.frame()
  
  wh.group = min(which(colnames(new.basics)=='group'))
  colnames(new.basics)[wh.group]=group.col
  
  if (value.col=='Hb') {
    freq.limits = freq %>% group_by(group,Sex) %>% summarise(limit=min(limit),.groups='drop') %>% mutate(year=group)
    new.basics = new.basics %>%
      inner_join(freq.limits,join_by(year,Sex)) %>%
      mutate(theoretical.deferred.prop = pnorm(limit,mean,sd)) %>%
      dplyr::select(-limit)
  }
  
  if (!is.null(age.freq)) {
    year.totals = age.freq %>%
      group_by(year,Sex) %>%
      summarise(n2=sum(n),age=mean(age),.groups='drop')
    
    age2 = age.freq %>%
      mutate(group = cut(age,age.breaks,labels=age.labels)) %>%
      group_by(year,Sex,group) %>%
      summarise(n=sum(n),.groups='drop') %>%
      inner_join(year.totals,join_by(year,Sex)) %>%
      mutate(prop=n/n2) %>% # ,name=paste(year,Sex)) %>%
      pivot_wider(id_cols=c('year','Sex'),names_from='group',values_from='prop')


    new.basics = new.basics %>%
      inner_join(age2,join_by(year,Sex)) %>%
      inner_join(year.totals[,c('year','Sex','age')],join_by(year,Sex))
  }

  
  return(new.basics)
}
```

```{r county-specific-basics}
getCountryBasics = function(hb.freq,age.freq,cf=1) {
  # hb.freq = data$hb.freq; age.freq=data$age.freq
  freq0=hb.freq[hb.freq$data.set=='donation0',]
  age0 = age.freq[age.freq$data.set=='donation0',]
  rv=by(freq0,freq0[,c('Sex','year')],function(x) unlist(rectifyDistribution(data0=NULL,freq=x,plot=FALSE)$params))
  res=array2DF(rv)
  res$year=as.integer(res$year)
  
  # Basic statistics for the unrectified hb distributions
  basics.all = getStats(freq0,age.freq = age0)
  
  # Basic statistics joined with results from the rectified distributions
  basics = inner_join(basics.all,res,
                      join_by(year,Sex),suffix=c('','.rectified'))
  
  if (cf==1) 
    return(basics)
  colnames(basics)
  
  convert.cols = grep('^mean|sd',colnames(basics))
  basics[,convert.cols] = basics[,convert.cols]*cf
  
  convert.cols.sq = grep('^var',colnames(basics))
  basics[,convert.cols.sq] = basics[,convert.cols.sq]*cf^2
  
  return(basics)
}
```


```{r basic-data-processing}
freq0=hb.freq[hb.freq$data.set=='donation0',]

# Rectify the male distribution from all years as an example
rectifyDistribution(data0=NULL,freq=freq0[freq0$Sex=='Male',])

rv=by(freq0,freq0[,c('Sex','year')],function(x) unlist(rectifyDistribution(data0=NULL,freq=x,plot=FALSE)$params))
res=array2DF(rv)
res$year=as.integer(res$year)

# Basic statistics for the unrectified hb distributions
basics.all = getStats(hb.freq,age.freq = age.freq)

# Basic statistics joined with results from the rectified distributions
basics = inner_join(basics.all,res,
                    join_by(year,Sex),suffix=c('','.rectified'))
```

``` {r plotting}
# means and such
plotColBySex(basics,'mean','mean (actual data)')
plotColBySex(basics,'mean.rectified','theoretical mean (rectified)')
plotColBySex(basics,'sd','sd (actual data)')
plotColBySex(basics,'sd.rectified','theoretical sd (rectified)')

# deferral rates
plotColBySex(basics,'deferred.prop','actual percentage deferred')
plotColBySex(basics,'theoretical.deferred.prop','theoretical proportion deferred (unrectified distribution)')
plotColBySex(basics,'deferred.prop.rectified','theoretical proportion deferred (rectified distribution)')
# plotColBySex(basics,'dperc.lm','adjusted using estimated mean differences')
```

``` {r new-lm-approach}
newLMapproach = function(basics) {
  # This block relies on basics, so need to make sure it can be computed with the summary data
  # Plus must add countries overall
  # This all seems to be based on basics
  par(mar=c(2,4,0,0))
  data=basics[basics$Sex=='Female',]
  data.female=data
  mean.female=mean(data$deferred.prop)
  plot(deferred.prop~mean,data=data,ylim=c(0,0.15),xlim=c(130,160),col='red')
  m=lm(deferred.prop~mean,data=data)
  m1=lm(deferred.prop~mean+sd,data=data)
  sm=summary(m)
  sm1=summary(m1)
  sm1.female=sm1
  
  m2.female=NULL
  if ('young' %in% colnames(data)) {
    m2.female=lm(deferred.prop~mean+sd+young+old,data=data)
  }

    abline(a=sm$coeff[1,1],b=sm$coeff[2,1],col='red',lty='dashed',lwd=2)
  text(x=param$cutoff.male,y=0.03,labels=paste0('b=',round(sm1$coeff[2,1],3),'\np=',round(sm1$coeff[2,4],3),'\nR',intToUtf8(178),'=',round(sm1$r.squared,3)),adj=c(0))
  data=basics[basics$Sex=='Male',]
  m=lm(deferred.prop~mean,data=data)
  m1=lm(deferred.prop~mean+sd,data=data)
  sm=summary(m)
  sm1=summary(m1)
  abline(a=sm$coeff[1,1],b=sm$coeff[2,1],col='blue',lty='dashed',lwd=2)
  points(data$mean,data$deferred.prop,col='blue')
  text(x=155,y=0.06,labels=paste0('b=',round(sm1$coeff[2,1],3),'\np=',round(sm1$coeff[2,4],3),'\nR',intToUtf8(178),'=',round(sm1$r.squared,3)),adj=c(0))
  plot(data$year,sm1.female$residuals,col='red',ylim=c(-0.05,0.05),type='l',ylab='residuals',lwd=3)
  lines(data$year,sm1$residuals,col='blue',lwd=3)
  mean.male=mean(data$deferred.prop)
  lines(data$year,data.female$deferred.prop-mean.female,col='red',lty='dotted',lwd=2)
  lines(data$year,data$deferred.prop-mean.male,col='blue',lty='dotted',lwd=2)
  
  year=data$year
  residual=data.female$deferred.prop-mean.female
  m=lm(residual~year)
  summary(m)
  
  table(data$Sex)
  m1=lm(deferred.prop~mean+sd+age,data=data)
  summary(m1)
  
  m1=lm(deferred.prop~mean+sd+age,data=data.female)
  summary(m1)

  m2.male=NULL
  if ('young' %in% colnames(data)) {
    m2.male=lm(deferred.prop~mean+sd+young+old,data=data)
  }
  
  return(list(m.female=m2.female,m.male=m2.male))
}
```

```{r experimental-results}
# Do the young and old age proportions explain deferral rates:
# They do, but only for men 
rv=newLMapproach(basics)
# summary(rv$m.female)
# summary(rv$m.male)

plotColBySex(basics,'old','proportion of 40+ yrs of age')
```


```{r statistics-by-age}
agedist = getStats(age.freq %>% filter(data.set=='all'),value.col='mean.hb',group.col='age',n.filter=100)
plotColBySex(agedist,'mean',xvar = 'age',ylab='')

plotColBySex(agedist,'n',xvar = 'age',ylab='')
plotColBySex(agedist,'mean',xvar = 'age',ylab='')
plotColBySex(agedist,'sd',xvar = 'age',ylab='')

plotColBySex(basics,'age',xvar='year',ylab='age by year')
```


